Surface-driven registration method for the structure-informed segmentation of diffusion MR images\tnoteref{t1}

### 2016-03-02 Publications/2015-NeuroImage/02-abstract.tex
remove comments from text

% -*- root: 00-main.tex -*-
Current methods for processing \gls*{dmri} to map the connectivity of the human brain
  require precise delineations of anatomical structures.
This requirement has been approached \revcomment[R\#3-C.33]{by} either segmenting the data in
  native \gls*{dmri} space or mapping the structural information from \gls*{t1} images.
The characteristic features of diffusion data in terms of \acrlong*{snr}, resolution, as well
  as the geometrical distortions caused by the inhomogeneity of magnetic susceptibility
  across tissues hinder both solutions.
Unifying the two approaches, we propose \regseg{}, a surface-to-volume nonlinear
  registration method that segments homogeneous regions within multivariate images by mapping
  a set of nested reference-surfaces.
Accurate surfaces are extracted from a \gls*{t1} image of the subject, using as target image
  the bivariate volume comprehending the \gls*{fa} and the \gls*{adc} maps derived from the
  \gls*{dmri} dataset.
We first verify the accuracy of \regseg{} on a general context using digital phantoms.
Then we establish an evaluation framework using undistorted \gls*{dmri} data from the \gls*{hcp}
  and known deformations derived from real inhomogeneity fieldmaps.
We analyze the performance of \regseg{} computing the misregistration error of the surfaces estimated
  after being mapped with \regseg{} onto 16 datasets from the \gls*{hcp}.
The distribution of errors shows a 95\% CI of 0.56--0.66 mm, that is below the \gls*{dmri}
  resolution (1.25 mm, isotropic).
Finally, we cross-compare the proposed tool against a nonlinear \lowb{}-to-T2w registration
  method, thereby obtaining a significantly lower misregistration error with \regseg{}.
Therefore, we demonstrate that \regseg{} allows the accurate mapping of structural information
  in \gls*{dmri} space, enabling the application of new structure-informed techniques in
  the connectome extraction.

>>>>

% -*- root: 00-main.tex -*-
Current methods for processing \gls*{dmri} to map the connectivity of the human brain
  require precise delineations of anatomical structures.
This requirement has been approached by either segmenting the data in
  native \gls*{dmri} space or mapping the structural information from \gls*{t1} images.
The characteristic features of diffusion data in terms of \acrlong*{snr}, resolution, as well
  as the geometrical distortions caused by the inhomogeneity of magnetic susceptibility
  across tissues hinder both solutions.
Unifying the two approaches, we propose \regseg{}, a surface-to-volume nonlinear
  registration method that segments homogeneous regions within multivariate images by mapping
  a set of nested reference-surfaces.
Accurate surfaces are extracted from a \gls*{t1} image of the subject, using as target image
  the bivariate volume comprehending the \gls*{fa} and the \gls*{adc} maps derived from the
  \gls*{dmri} dataset.
We first verify the accuracy of \regseg{} on a general context using digital phantoms 
  distorted with synthetic and random deformations.
Then we establish an evaluation framework using undistorted \gls*{dmri} data from the \gls*{hcp}
  and realistic deformations derived from 
  the inhomogeneity fieldmap corresponding to each subject.
We analyze the performance of \regseg{} computing the misregistration error of the surfaces estimated
  after being mapped with \regseg{} onto 16 datasets from the \gls*{hcp}.
The distribution of errors shows a 95\% CI of 0.56--0.66 mm, that is below the \gls*{dmri}
  resolution (1.25 mm, isotropic).
Finally, we cross-compare the proposed tool against a nonlinear \lowb{}-to-T2w registration
  method, thereby obtaining a significantly lower misregistration error with \regseg{}.
The accurate mapping of structural information in \gls*{dmri} space 
  is fundamental to increase the reliability of network building
  in connectivity analyses, and to improve the performance of the emerging structure-informed
  techniques for \gls*{dmri} data processing.

### 2016-03-02 Publications/2015-NeuroImage/03-introduction.tex
remove comments from text

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:regseg-intro}
\Acrlong*{dmri} enables the mapping of microstructure \citep{basser_microstructural_1996}
  and connectivity \citep{craddock_imaging_2013} of the human brain \emph{in-vivo}.
It is generally acquired using \gls*{epi} schemes, since they are very fast at
  scanning a large sequence of images called \glspl*{dwi}.
Each \gls*{dwi} is sensitized with a gradient to probe proton diffusion in a certain
  orientation.
Subsequent processing involves describing the local microstructure with one of the available
  models, which range from the early \gls*{dti} proposed by \cite{basser_microstructural_1996}
  to current models such as \revcomment[R\#3-C.10]{\gls*{amico}}.
The microstructural map is then used to draw the preferential orientations of diffusion
  across the brain using tractography \citep{mori_threedimensional_1999}.
Finally, a graph representing the corresponding structural network is built using
  the regions of a cortical parcellation as nodes and the fiber paths found by
  tractography as edges \citep{hagmann_mapping_2008}.
The methodologies to solve reconstruction, tractography and network building
  require the delineation of the anatomy in the \gls*{dmri} space.
Moreover, current trends on reconstruction \citep{jeurissen_multitissue_2014} and
  tractography \citep{smith_anatomicallyconstrained_2012} are increasingly using
  structural information to improve the microstructural mapping and fiber-tracking.

Possibly, the earliest structural information incorporated to aid \gls*{dmri} processing
  is the \gls*{wm} mask used as a termination criteria for tractography.
The standardized procedure to obtain this mask was thresholding the \gls*{fa} map.
However, the mask and subsequent analyses are highly dependent on the 
  \revcomment[R\#3-C.33]{threshold that is chosen} \citep{taoka_fractional_2009}.
To overcome the unreliability of \gls*{fa} thresholding, and to broaden
  \gls*{wm} segmentation to brain tissue segmentation, a large number of
  methods have been proposed using \glspl*{dwi}, the \lowb{}, and \gls*{dti}-derived
  scalar maps such as \gls*{fa}, \gls*{adc} and others \citep{zhukov_level_2003,
  rousson_level_2004,jonasson_segmentation_2005,hadjiprocopis_unbiased_2005,liu_brain_2007,
  lu_segmentation_2008,han_experimental_2009}.
However, the precise segmentation of \gls*{dmri} is difficult for several reasons.
First, \gls{dmri} images have a resolution that is much lower
  than that of the imaged microstructural features.
Therefore, voxels located in structural discontinuities are affected by partial
  voluming of the signal sources.
Second, the extremely low \gls*{snr} and the high dimensionality of the \glspl*{dwi} prevent
  their direct use in segmentation.
Third, the low contrast between \gls*{gm} and \gls*{wm} in the \lowb{} volume also makes
  it unsuitable for brain tissue segmentation.

@@

Here, we present a segmentation and surface-to-volume registration method
  called \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
\revcomment[authors]{We apply \regseg{} to segment \gls*{dmri} data by mapping
  a} set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) to
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol \revcomment[R\#1-C.2]{in order to estimate the deformation field};
  2) \revcomment[R\#1-C.2]{depending on the application, the resampling of the \glspl*{dwi}
    introduced by the correction method can be avoided, either by performing the posterior
    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
    itself}; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
Then, we evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
\revcomment[authors]{This approach is the first step of the} \gls*{t2b} correction
  methods \citep{kybic_unwarping_2000}.
\revcomment[authors]{We reproduce the settings and implementation of a} widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:regseg-intro}
\Acrlong*{dmri} enables the mapping of microstructure \citep{basser_microstructural_1996}
  and connectivity \citep{craddock_imaging_2013} of the human brain \emph{in-vivo}.
It is generally acquired using \gls*{epi} schemes, since they are very fast at
  scanning a large sequence of images called \glspl*{dwi}.
Each \gls*{dwi} is sensitized with a gradient to probe proton diffusion in a certain
  orientation.
Subsequent processing involves describing the local microstructure with one of the available
  models, which range from the early \gls*{dti} proposed by \cite{basser_microstructural_1996}
  to current models such as \gls*{amico}.
The microstructural map is then used to draw the preferential orientations of diffusion
  across the brain using tractography \citep{mori_threedimensional_1999}.
Finally, a graph representing the corresponding structural network is built using
  the regions of a cortical parcellation as nodes and the fiber paths found by
  tractography as edges \citep{hagmann_mapping_2008}.
The methodologies to solve reconstruction, tractography and network building
  require the delineation of the anatomy in the \gls*{dmri} space.
Moreover, current trends on reconstruction \citep{jeurissen_multitissue_2014} and
  tractography \citep{smith_anatomicallyconstrained_2012} are increasingly using
  structural information to improve the microstructural mapping and fiber-tracking.

Possibly, the earliest structural information incorporated to aid \gls*{dmri} processing
  is the \gls*{wm} mask used as a termination criteria for tractography.
The standardized procedure to obtain this mask was thresholding the \gls*{fa} map.
However, the mask and subsequent analyses are highly dependent on the 
  threshold that is chosen \citep{taoka_fractional_2009}.
To overcome the unreliability of \gls*{fa} thresholding, and to broaden
  \gls*{wm} segmentation to brain tissue segmentation, a large number of
  methods have been proposed using \glspl*{dwi}, the \lowb{}, and \gls*{dti}-derived
  scalar maps such as \gls*{fa}, \gls*{adc} and others \citep{zhukov_level_2003,
  rousson_level_2004,jonasson_segmentation_2005,hadjiprocopis_unbiased_2005,liu_brain_2007,
  lu_segmentation_2008,han_experimental_2009}.
However, the precise segmentation of \gls*{dmri} is difficult for several reasons.
First, \gls{dmri} images have a resolution that is much lower
  than that of the imaged microstructural features.
Therefore, voxels located in structural discontinuities are affected by partial
  voluming of the signal sources.
Second, the extremely low \gls*{snr} and the high dimensionality of the \glspl*{dwi} prevent
  their direct use in segmentation.
Third, the low contrast between \gls*{gm} and \gls*{wm} in the \lowb{} volume also makes
  it unsuitable for brain tissue segmentation.

@@

Here, we present a segmentation and surface-to-volume registration method
  called \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
We apply \regseg{} to segment \gls*{dmri} data by mapping
  a set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) to
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol in order to estimate the deformation field;
  2) depending on the application, the resampling of the \glspl*{dwi}
    introduced by the correction method can be avoided, either by performing the posterior
    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
    itself; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
Then, we evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
This approach is the first step of the \gls*{t2b} correction
  methods \citep{kybic_unwarping_2000}.
We reproduce the settings and implementation of a widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.
